---
title: 2025 WIT Reporting
date: 2025-11-14
tags: []
format:
  commonmark:
    variant: +yaml_metadata_block
    df-print: kable
execute:
  echo: false
  message: false
  warning: false
  class-output: 'r'
---

For FY 25 this is an explanation of WIT reporting and sources.

```{r}
library(tidyverse)
library(arcgislayers)
library(sf)
library(sjnftools)

get_treat_fy <- function(url, fy, clip_bd) {
  con <- arc_open(url)

  where_st <- paste("FISCAL_YEAR_COMPLETED=", fy)

  fy_2025 <- con |>
    arc_select(where = where_st)

  crd_2025 <- fy_2025 |>
    st_make_valid() |>
    st_transform(st_crs(clip_bd)) |>
    st_crop(clip_bd)

  crd_2025 |>
    select(
      TREATMENT_NAME,
      ACTIVITY_UNIT_NAME,
      GIS_ACRES,
      ACTIVITY,
      NEPA_DOC_NAME,
      IMPLEMENTATION_PROJECT,
      IMPLEMENTATION_PROJECT_NBR,
      IMPLEMENTATION_PROJECT_TYPE,
      EQUIPMENT,
      FUND_CODE,
      METHOD
    )
}
```

## RX Fires on CRD
RX fire activities were pulled from the FACTS database.
```{r}
# rx_fire <- get_treat_fy(
#   url = "https://apps.fs.usda.gov/arcx/rest/services/EDW/EDW_HazardousFuelsTreatments_01/MapServer/3",
#   fy = 2025,
#   clip_bd = crd_bd
# )
#
# rx_fire |>
#   st_write("content/blog/2025/2025_wit_reporting/cache/rx_fire.gpkg")

rx_fire <- st_read("./cache/rx_fire.gpkg", quiet = T)


rx_fire |>
  st_drop_geometry() |>
  as_tibble()
```


## Mechanical Treatments
Mechanical Treatments were also pulled from FACTS.
```{r}
# mech_treat_crd <- get_treat_fy(
#   url = "https://apps.fs.usda.gov/arcx/rest/services/EDW/EDW_HazardousFuelsTreatments_01/MapServer/5",
#   fy = 2025,
#   clip_bd = crd_bd
# )
#
# st_write(
#   mech_treat_crd,
#   "content/blog/2025/2025_wit_reporting/cache/mx_treat.gpkg"
# )

mech_treat_crd <- st_read("./cache/mx_treat.gpkg", quiet = T) |>
  filter(ACTIVITY == "Improvement Cut")


sum_imp_cut <- mech_treat_crd |>
  pull(GIS_ACRES) |>
  sum() |>
  round(0)

mech_treat_crd |>
  st_drop_geometry() |>
  as_tibble()
```

We completed `r sum_imp_cut` acres of mechanical treatment in 2025.  These acres will may not count toward WIT because we only consider treatments in ponderosa pine that seek to create openings, clumps and groups as wildlife treatments.  Treatments that strictly seek to meet a basal area, while effective for fuels mitigation, do very little to improve or maintain habitat for wildlife and therefore are not counted.

We have one more treatment, Turkey Springs, that has not been added to FACTS yet and I was not able to pull into this report.

## Weed Treatments

```{r}
## library(arcgisbinding)
## arc.check_product()
##
## nox_weeds_treat<-arc.open("T:\\FS\\Reference\\GIS\\r02_sjnf\\LayerFile\\SanJuan_Library\\NoxiousWeeds\\TESP_InvasiveSpecies_NRM\\Invasive Species Accomplished Activities_RSW.lyr")  |>
## 	arc.select() |>
## 	arc.data2sf()
##
## sjnf_nox_weeds<-nox_weeds_treat |>
## 	st_make_valid() |>
## 	st_transform(st_crs(sjnf_bd)) |>
## 	st_intersection(sjnf_bd)  |>
## 	filter(FISCAL_YEAR_ACCOMPLISHED>2020)
##
##
## st_write(sjnf_nox_weeds,"content/blog/2025/2025_wit_reporting/cache/nox_weeds_sjnf_bd.gpkg")

# sjnf_nox_weeds<-st_read("content/blog/2025/2025_wit_reporting/cache/nox_weeds_sjnf_bd.gpkg")

sjnf_nox_weeds <- st_read("./cache/nox_weeds_sjnf_bd.gpkg", quiet = T)

sjnf_nox_weeds |>
  st_drop_geometry() |>
  count(FISCAL_YEAR_COMPLETED)
```







