---
title: FIA Ponderosa Pine Analysis
date: 2025-09-04
author: Mike Schmidt
tags: ["Ecology", "Ponderosa Pine", "FIA"]
format:
  commonmark:
    variant: +yaml_metadata_block
    df-print: tibble
execute:
  echo: false
  message: false
  warning: false
  class-output: 'r'
---

Ponderosa pine and warm-dry mixed conifer historically experienced low severity frequent fires.  Fire has largely been absence from these systems. I've been trying to learn about these systems in hope of better managing them for wildlife.  This note will review several papers with management recommendations for ponderosa pine and will also analyze ponderosa pine on the San Juan using FIA data.


```{r}
library(tidyverse)
library(sf)
library(sjnftools)
library(rFIA)
library(ggrain)
library(MetBrewer)

theme_set(theme_minimal())
```

```{r}
co <- readFIA(
  "C:\\Users\\MichaelSchmidt2\\OneDrive - USDA\\code\\r\\test\\fia\\download"
)

sjnf <- clipFIA(
  co,
  mostRecent = FALSE,
  mask = sjnf_bd
)

sjnf_species <- tpa(
  sjnf,
  bySpecies = T,
  grpBy = "PLOT"
) |>
  janitor::clean_names()

sjnf_species
```


```{r}
sjnf_plots <- sjnf$PLOT |>
  as_tibble() |>
  janitor::clean_names() |>
  group_by(lon, lat) |>
  summarize(
    n = n(),
    max_year = max(invyr, na.rm = T),
    min_year = min(invyr, na.rm = T)
  ) |>
  ungroup() |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

mapview::mapview(sjnf_plots)
```


```{r}
sjnf_species |>
  filter(common_name == "ponderosa pine") |>
  ggplot(aes(as.factor(year), baa, fill = as.factor(year))) +
  geom_rain(
    violin.args.pos = list(
      side = "r",
      width = 1.7,
      position = position_nudge(x = 0.2)
    )
  ) +
  coord_flip() +
  scale_fill_manual(values = met.brewer("Renoir", 17))
```






