---
title:  "Weminuche Bighorn Sheep (RBS-20)"
date: 2023-04-28
tags: ["Bighorn Sheep"]
format:
  commonmark:
    variant: +yaml_metadata_block
    df-print: kable
execute:
  echo: false
  message: false
  warning: false
  class-output: 'r'
---

## General
* Population Estimate (in DAU Plan 2012): 460 (200 in S-15, 135 in S-16, and 125 in S28) (CPW 2012)
* Population Estimate 2022: 395, a decrease of 17% from 2012 (180 in S15, 135 in S16, 80 in S28). The largest decrease was in S-28 (Vallecito), a decrease of 36%.
* Population Estimate in 2023: Decreased!!
  * S28 - 50 (was 80)
  * S16 - 100 (was 135)
  * S15 - 140 (was 180)
  * Total - 290 (was 395 [26% reduction])
* Statewide 6,900.

## Conservation Status Notes
* [CPW Statewide Action Plan](https://cpw.state.co.us/Documents/WildlifeSpecies/SWAP/CO_SWAP_FULLVERSION.pdf): Tier 2
* Forest Service: [Sensitive](https://www.fs.usda.gov/Internet/FSE_DOCUMENTS/stelprdb5318798.pdf)
* BLM Colorado: Sensitive
* CPW Satus: **Tier 1 Herd** (A Primary Population). Tier 1 herds: "are regarded as those large (i.e., ≥100 animals for ≥90% of the years since 1986), native populations comprised of one or more interconnected herds (in, or to be designated into, GMUs) that have received few (i.e., ≤50 animals total) if any supplemental releases of Rocky Mountain bighorn sheep in the past. Tier 1 populations should be given the highest priority for inventory, habitat protection and improvement, disease prevention, and research." (CPW 2021)
* [Nature Serve: Worldwide](https://explorer.natureserve.org/Taxon/ELEMENT_GLOBAL.2.102557/Ovis_canadensis) = G4 (Apparently Secure), Colorado = S4 (Apprently Secure)

## History
Excerpts from CPW DAU Plan:

* 1911 - "Forest Ranger E. E. Chapson, of Pagosa Springs, says that a good many sheep are killed by snowslides in the San Juan Mountains north of that point."
* “In 1962 Moser documented locations occupied by bighorn sheep representing a large geographical area within the San Juan Mountains. This included Cimarron Peak, South River Peak, Weminuche Creek, Grenadier Range, Sunshine Peak, and Engineer Mountain”
* “By 1973 the range of bighorn sheep had decreased significantly from historic occurrence. Summer range consisted of three main areas. The first was the country north of Wolf Creek Pass to Table Mountain and west to Beaver Creek. The second area included the headwaters of Weminuche, Cimarrona, Hossick, and Williams Creeks. And the third was the region between the Los Pinos River and Vallecito Creek north to Rock Creek (Bear and Jones 1973). The western areas, past Vallecito Creek, were no longer presented as occupied habitat.”

```{r }
library(tidyverse)
library(sjnftools)
library(sf)

theme_set(theme_void(base_size = 9))
```

```{r}

bhov <- cpw_sam$bighorn_summer_range |>
  st_transform(st_crs(crd_bd)) |>
  st_crop(crd_bd)

allot_cl <- sjnf_pastures |>
  st_make_valid() |>
  filter(districtname == "Columbine Ranger District") |>
  group_by(allotment_name, pasture_status) |>
  summarize() |>
  ungroup() |>
  filter(
    allotment_name %in%
      c(
        "BURNT TIMBER",
        "CANYON CREEK",
        "CAVE BASIN",
        "ENDLICH MESA",
        "SPRING GULCH",
        "FALL CREEK",
        "FLINT CREEK",
        "JOHNSON CREEK",
        "LEVIATHAN",
        "PINE RIVER",
        "ROCK CREEK",
        "TANK CREEK",
        "VIRGINIA GULCH"
      )
  ) |>
  mutate(
    pasture_auth = case_when(
      allotment_name %in%
        c(
          "BURNT TIMBER",
          "CANYON CREEK",
          "ENDLICH MESA",
          "SPRING GULCH",
          "TANK CREEK",
          "VIRGINIA GULCH"
        ) ~
        "EIS Proposed Sheep Allot",
      TRUE ~ "EIS Proposed Vacant Allotment"
    )
  )

ggplot() +
  geom_sf(data = crd_bd, fill = "grey70") +
  geom_sf(
    data = allot_cl,
    aes(fill = pasture_auth),
    color = "grey30",
    linewidth = 0.5
  ) +
  scale_fill_manual(values = c("#27AE60", "#FF5733", "#FFC300")) +
  geom_sf_label(
    data = allot_cl,
    aes(label = allotment_name),
    size = 1.5,
    show.legend = FALSE
  ) +
  geom_sf(data = bhov, aes(fill = full_name), alpha = 0.5, linewidth = 1) +
  labs(
    title = "Grazing allotments and Bighorn Sheep Overall Range",
    subtitle = "Allotment status as proposed in Draft Weminuche Landscape Grazing Analysis 2016",
    color = "Pasture Proposed Status"
  ) +
  theme(
    legend.position = c(0.15, 0.15)
  )

active_allot <- allot_cl |>
  filter(
    allotment_name %in%
      c(
        "BURNT TIMBER",
        "SPRING GULCH",
        "TANK CREEK",
        "VIRGINIA GULCH"
      )
  )
```


```{r}
obs <- read_csv("data/Observations_3_15_2023.csv") |>
  st_as_sf(coords = c("longitude", "latitude"), crs = st_crs(4326)) |>
  st_transform(st_crs(crd_bd)) |>
  st_crop(crd_bd) |>
  mutate(
    label = "Bighorn Sheep Observation"
  )

ggplot() +
  geom_sf(data = crd_bd, fill = "grey70") +
  geom_sf(
    data = allot_cl,
    aes(fill = pasture_auth),
    color = "grey30",
    linewidth = 0.5
  ) +
  scale_fill_manual(
    values = c("#FF5733", "#FFC300"),
    name = "EIS Allotment Status",
    guide = guide_legend(override.aes = list(shape = NA))
  ) +
  geom_sf_label(
    data = allot_cl,
    aes(label = allotment_name),
    size = 1.5,
    show.legend = FALSE
  ) +
  geom_sf(
    data = obs,
    aes(color = label),
    shape = 15,
    size = 4,
    alpha = 0.7,
    show.legend = "point"
  ) +
  scale_color_manual(values = "#222222") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(
    title = "MSI Citizen Science Bighorn Observations",
    subtitle = "Columbine Ranger District. Allotments status are draft, proposed in the 2016 Weminuche Grazing EIS",
    caption = "Data: iNaturalist",
    color = "MSI Provided Data"
  ) +
  theme(
    legend.position = c(0.15, 0.15)
  )
```

## Collaring and Disease Study

### Collaring Effort

## Todo
[ ] - Need to look in more detail the observations that overlap with non-vacant allotments.
[ ] - add active allotments in Silverton.

