---
title: What Trails are closed outside of Durango
date: 2025-12-08
tags: ["Trails"]
format:
  commonmark:
    df-print: kable
    variant: +yaml_metadata_block
    prefer-html: true
execute:
  echo: false
  message: false
  warning: false
  class-output: "r"
---

```{r}
library(tidyverse)
library(sf)
library(sjnftools)
library(arcgislayers)
library(units)
library(mapview)
```

This analysis is a look at trails near Durango, Colorado. I wanted to know what proportion of trails within 5 miles of Durango are closed during the winter. I was starting to feel like the majority of trails have wildlife closures on them in the winter and that as wildlfie advocates we may want to stop closing areas because the public was running out of recreational opportunities near durango. This analsyis will show that is not the case.

```{r}

# Define coordinates (Easting and Northing in meters)
easting <- 244640.55967237
northing <- 4129299.7118673

# Create a data frame with coordinates
coords <- data.frame(x = easting, y = northing)

# Convert to sf object with specified CRS (EPSG:26913 = UTM Zone 13N NAD83)
durango <- st_as_sf(coords, coords = c("x", "y"), crs = 26913)
durango_buffer <- durango |>
  st_buffer(1600 * 5)
```

```{r}
near_durango <- co_ownership |>
  st_intersection(durango_buffer) |>
  filter(adm_name != "PRIVATE") |>
  mutate(
    winter_closure = case_when(
      OBJECTID %in% c(10538, 12712, 7239, 12840, 8012) ~ "Yes",
      TRUE ~ "No"
    )
  )
```

```{r}
con_open <- arc_open(
  "https://services5.arcgis.com/ttNGmDvKQA7oeDQ3/arcgis/rest/services/CPWAdminData/FeatureServer/15/query?outFields=*&where=1%3D1"
)

trails_near_durango <- con_open |>
  arc_select(filter_geom = st_bbox(durango_buffer), crs = 26913)

trails_near_durango_cl <- trails_near_durango |>
  st_intersection(near_durango) |>
  mutate(
    length = st_length(geometry),
    length_miles = set_units(length, "miles")
  )


trails_near_durango_cl |>
  as_tibble() |>
  group_by(winter_closure) |>
  summarize(
    length = sum(length),
    length_miles = sum(length_miles)
  ) |>
  ungroup() |>
  mutate(
    perc = length_miles / sum(length_miles)
  )
```

## Conclusion

Only about 30% of trails in Durango have winter closures on them. This is fairly crude estimate given that the trail dataset I used does not include trails on private property in Durango.

Need to add a map here:

```{r}

```
